---
title: "02_tf_activity"
author: "Tabea M. Soelter"
date: '2023-09-27'
output: html_document
---
**Transcription factor activity analysis using decoupleR**

__Goal__: Infer TF activity from pseudo-bulked DGE analyses. The TFs I am investigating are signaling mediators of LRTs that overlap across independent datasets.

__Reproducibility__: 
* GitHub: lasseignelab/230313_TS_CCCinHumanAD
* Docker: tsoelter/rstudio_ccc_ad
    * Version: 1.0.5
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 65GB per CPU

__Data__:
*DESeq2 outputs*
* Name: Inhibitory Neurons_group_id_AD_vs_CTRL_all_genes.csv
* Location: results/intermediate_outputs/geo/pseudobulk/ & results/intermediate_outputs/gse/pseudobulk/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * DEGs from DESeq2
  * Signaling mediators from MultiNicheNet
* Download prior
  * CollecTRI
* Prepare for TF activity analysis
  * Create 1 matrix which has DEGs from both datasets
* Differential TF activity analysis using Multivariate Linear Model (MLM)
* Save plot and output
  
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Setting locale
options(encoding = "UTF-8")

# Load packages
suppressPackageStartupMessages({
  library(decoupleR)
  library(OmnipathR)
  library(tidyverse)
  library(ComplexHeatmap)
  library(circlize)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCinAD_tms.R"))
```

# Load data
```{r}
# signaling mediators in inhibitory neurons
in_mediators <- c(
  "GSK3B", "IRAK1", "CD44", "ESR1", "JUN", "MYC", "SRC",
  "STAT3", "TLR2", "TP53", "AKT1", "CTNNB1", "ERBB2", "DAPK1", "RAC1", "LRP1",
  "MAPK8", "APP", "LCK", "PTK2", "STAT1", "STAT2", "PTEN", "CCNB1",
  "CCND1", "IFIT3", "NFIL3", "ERBB3", "CEBPD", "FST", "GFRA1", "ANGPTL4"
)

# DEGs in inhibitory neurons from DESeq2
geo_degs_in <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "pseudobulk",
    "Inhibitory Neurons_group_id_AD_vs_CTRL_all_genes_final.csv"
  )
)

gse_degs_in <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "gse",
    "pseudobulk",
    "Inhibitory Neurons_group_id_AD_vs_CTRL_all_genes_final.csv"
  )
)
```

# Access prior
```{r}
collectri <- get_collectri(organism = "human", split_complexes = FALSE)
# Downloaded 64958 interactions
# Downloaded 232 interactions
```

# Prepare data
```{r}
## dataset 1
df1 <- geo_degs_in %>%
  select(stat, gene) %>%
  pivot_wider(
    names_from = "gene",
    values_from = "stat"
  ) %>%
  as.data.frame()

# make dataset name the rowname
rownames(df1) <- "geo"

# replace NAs
df1[is.na(df1)] <- 0

# turn into matrix
mat1 <- as.matrix(df1)

## dataset 2
df2 <- gse_degs_in %>%
  select(stat, gene) %>%
  pivot_wider(
    names_from = "gene",
    values_from = "stat"
  ) %>%
  as.data.frame()

# make dataset name the rowname
rownames(df2) <- "gse"

# replace NAs
df2[is.na(df2)] <- 0

# turn into a matrix
mat2 <- as.matrix(df2)

## Combine datasets
df <- merge(mat1, mat2, all = TRUE)

# make sure no NAs remain, replace with 0 again
df[is.na(df)] <- 0

# turn into matrix
df <- as.matrix(df)

rownames(df) <- c("geo", "gse")
```

# TF activity analysis
* We use genes that are differentially expressed (as determined by DESeq2), therefore any TF activity observed, is different between AD and CTRL.
  * Since CTRL was our base, any changes are to be interpreted with respect to AD
* We use the Multivariate Linear Model (MLM) from decoupleR
```{r}
# TF activity analysis
tf_acts_in <- run_mlm(
  mat = t(df),
  network = collectri
)

# filter by mediators
tf_acts_filt <- tf_acts_in %>%
  filter(source %in% in_mediators)

dim(tf_acts_filt) # 20 5
```

# Plot heatmap
```{r}
# prepare matrix for plotting
mat <- tf_acts_filt %>%
  select(source, condition, score) %>%
  pivot_wider(
    names_from = "source",
    values_from = "score"
  ) %>%
  column_to_rownames(var = "condition") %>%
  as.matrix()

# prepare top annotation
dataset <- c("geo", "gse")
cell_type <- c("Inhibitory Neurons", "Inhibitory Neurons")
meta <- cbind(dataset, cell_type) %>% as.data.frame()
rownames(meta) <- c("geo", "gse")

# Set annotation colors
anno_cols <- list(
  "cell_type" = c("Inhibitory Neurons" = "darkslategray3"),
  "dataset" = c(
    "geo" = "firebrick3",
    "gse" = "goldenrod3"
  )
)

# Make heatmap annotation (top)
anno <- HeatmapAnnotation(
  df = meta,
  show_annotation_name = FALSE,
  col = anno_cols
)

# Set heatmap colors and breaks
cols <- colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

tf_activity_heatmap <- Heatmap(t(mat),
  col = cols,
  top_annotation = anno,
  show_column_names = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  cluster_columns = TRUE,
  cluster_rows = TRUE
)

tf_activity_heatmap
```

# Save plot
```{r}
png(here("results", "final_outputs", "comparison", "tf_activity_heatmap_final.png"))
tf_activity_heatmap
dev.off()
```

# save TF activity outputs
```{r}
saveRDS(
  tf_acts_in,
  here("results",
       "final_outputs",
       "comparison",
       "differential_tf_activity_all_final.rds")
)

saveRDS(
  tf_acts_filt,
  here("results",
       "final_outputs",
       "comparison",
       "differential_tf_activity_filt_final.rds")
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.1.3 (2022-03-10)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so

locale:
[1] C

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2           styler_1.9.1          here_1.0.1            circlize_0.4.15      
 [5] ComplexHeatmap_2.10.0 lubridate_1.9.2       forcats_1.0.0         stringr_1.5.0        
 [9] dplyr_1.1.3           purrr_1.0.2           readr_2.1.4           tidyr_1.3.0          
[13] tibble_3.2.1          ggplot2_3.4.2         tidyverse_2.0.0       OmnipathR_3.9.8      
[17] decoupleR_2.7.1      

loaded via a namespace (and not attached):
 [1] matrixStats_1.0.0   bit64_4.0.5         doParallel_1.0.17   RColorBrewer_1.1-3 
 [5] progress_1.2.2      httr_1.4.7          rprojroot_2.0.3     R.cache_0.16.0     
 [9] tools_4.1.3         backports_1.4.1     utf8_1.2.3          R6_2.5.1           
[13] lazyeval_0.2.2      BiocGenerics_0.40.0 colorspace_2.1-0    GetoptLong_1.0.5   
[17] withr_2.5.0         processx_3.8.1      tidyselect_1.2.0    prettyunits_1.2.0  
[21] bit_4.0.5           curl_5.0.2          compiler_4.1.3      cli_3.6.1          
[25] rvest_1.0.3         xml2_1.3.5          desc_1.4.2          scales_1.2.1       
[29] checkmate_2.2.0     callr_3.7.3         rappdirs_0.3.3      digest_0.6.33      
[33] R.utils_2.12.2      rmarkdown_2.25      pkgconfig_2.0.3     htmltools_0.5.6    
[37] parallelly_1.36.0   fastmap_1.1.1       rlang_1.1.1         GlobalOptions_0.1.2
[41] readxl_1.4.3        rstudioapi_0.14     shape_1.4.6         generics_0.1.3     
[45] jsonlite_1.8.7      vroom_1.6.3         R.oo_1.25.0         magrittr_2.0.3     
[49] Matrix_1.5-4        Rcpp_1.0.11         munsell_0.5.0       S4Vectors_0.32.4   
[53] fansi_1.0.4         logger_0.2.2        lifecycle_1.0.3     R.methodsS3_1.8.2  
[57] stringi_1.7.12      yaml_2.3.7          parallel_4.1.3      crayon_1.5.2       
[61] lattice_0.20-45     hms_1.1.3           ps_1.7.5            knitr_1.44         
[65] pillar_1.9.0        igraph_1.5.1        rjson_0.2.21        codetools_0.2-18   
[69] stats4_4.1.3        glue_1.6.2          evaluate_0.21       rex_1.2.1          
[73] remotes_2.4.2       selectr_0.4-2       png_0.1-8           vctrs_0.6.3        
[77] tzdb_0.4.0          foreach_1.5.2       cellranger_1.1.0    gtable_0.3.3       
[81] clue_0.3-64         xfun_0.40           cyclocomp_1.1.0     later_1.3.1        
[85] iterators_1.0.14    IRanges_2.28.0      cluster_2.1.2       timechange_0.2.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 10 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "biological_activity",
  "02_tf_activity.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "biological_activity",
  "02_tf_activity.Rmd"
))
```

