---
title: "03_pathway_activity"
author: "Tabea M. Soelter"
date: '2023-10-03'
output: html_document
---
**Pathway activity analysis using decoupleR**

__Goal__: Infer pathway activity from pseudo-bulked DGE analyses in order to complement TF activity analyses performed previously.

__Reproducibility__: 
* GitHub: lasseignelab/230313_TS_CCCinHumanAD
* Docker: tsoelter/rstudio_ccc_ad
    * Version: 1.0.5
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 65GB per CPU

__Data__:
*DESeq2 outputs*
* Name: Inhibitory Neurons_group_id_AD_vs_CTRL_all_genes.csv
* Location: results/intermediate_outputs/geo/pseudobulk/ & results/intermediate_outputs/gse/pseudobulk/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * DEGs from DESeq2
* Download prior
  * PROGENy
* Prepare for pathway activity analysis
  * Create 1 matrix which has DEGs from both datasets
* Differential pathway activity analysis using Multivariate Linear Model (MLM)
* Save plot and output
  
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Setting locale
options(encoding = "UTF-8")

# Load packages
suppressPackageStartupMessages({
  library(decoupleR)
  library(OmnipathR)
  library(tidyverse)
  library(ComplexHeatmap)
  library(circlize)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCinAD_tms.R"))
```

# Load data
```{r}
# DEGs in inhibitory neurons from DESeq2
geo_degs_in <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "pseudobulk",
    "Inhibitory Neurons_group_id_AD_vs_CTRL_all_genes_final.csv"
  )
)

gse_degs_in <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "gse",
    "pseudobulk",
    "Inhibitory Neurons_group_id_AD_vs_CTRL_all_genes_final.csv"
  )
)
```

# Access prior
* Downloading PROGENy model
```{r}
net <- get_progeny(organism = "human", top = 500)

dim(net) # 7000 4
```

# Prepare data
```{r}
## dataset 1
df1 <- geo_degs_in %>%
  select(stat, gene) %>%
  pivot_wider(
    names_from = "gene",
    values_from = "stat"
  ) %>%
  as.data.frame()

# make dataset name the rowname
rownames(df1) <- "geo"

# replace NAs
df1[is.na(df1)] <- 0

# turn into matrix
mat1 <- as.matrix(df1)

## dataset 2
df2 <- gse_degs_in %>%
  select(stat, gene) %>%
  pivot_wider(
    names_from = "gene",
    values_from = "stat"
  ) %>%
  as.data.frame()

# make dataset name the rowname
rownames(df2) <- "gse"

# replace NAs
df2[is.na(df2)] <- 0

# turn into a matrix
mat2 <- as.matrix(df2)

## Combine datasets
df <- merge(mat1, mat2, all = TRUE)

# make sure no NAs remain, replace with 0 again
df[is.na(df)] <- 0

# turn into matrix
df <- as.matrix(df)

rownames(df) <- c("geo", "gse")
```

# Pathway activity analysis
* We use genes that are differentially expressed (as determined by DESeq2), therefore any pathway activity observed, is different between AD and CTRL.
  * Since CTRL was our base, any changes are to be interpreted with respect to AD
* We use the Multivariate Linear Model (MLM) from decoupleR
```{r}
# Pathway activity analysis
acts <- run_mlm(t(df),
  net,
  .source = "source",
  .target = "target",
  .mor = "weight",
  minsize = 5
)
```

# Plot heatmap
```{r}
# prepare matrix for plotting
mat <- acts %>%
  select(source, score, condition) %>%
  pivot_wider(
    names_from = source,
    values_from = score
  ) %>%
  column_to_rownames(var = "condition") %>%
  as.matrix()

# prepare top annotation
dataset <- c("geo", "gse")
cell_type <- c("Inhibitory Neurons", "Inhibitory Neurons")
meta <- cbind(dataset, cell_type) %>% as.data.frame()
rownames(meta) <- c("geo", "gse")

# Set annotation colors
anno_cols <- list(
  "cell_type" = c("Inhibitory Neurons" = "darkslategray3"),
  "dataset" = c(
    "geo" = "firebrick3",
    "gse" = "goldenrod3"
  )
)

# Make heatmap annotation (top)
anno <- HeatmapAnnotation(
  df = meta,
  show_annotation_name = FALSE,
  col = anno_cols
)

# Set heatmap colors and breaks
cols <- colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

pathway_activity_heatmap <- Heatmap(t(mat),
  col = cols,
  top_annotation = anno,
  show_column_names = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  cluster_columns = TRUE,
  cluster_rows = TRUE
)

pathway_activity_heatmap
```

# Save plot
```{r}
png(here(
  "results",
  "final_outputs",
  "comparison",
  "pathway_activity_heatmap_final.png"
))
pathway_activity_heatmap
dev.off()
```

# save pathway activity output
```{r}
saveRDS(
  acts,
  here(
    "results",
    "final_outputs",
    "comparison",
    "differential_pathway_activity_final.rds"
  )
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.1.3 (2022-03-10)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so

locale:
[1] C

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2           styler_1.9.1          here_1.0.1            circlize_0.4.15      
 [5] ComplexHeatmap_2.10.0 lubridate_1.9.2       forcats_1.0.0         stringr_1.5.0        
 [9] dplyr_1.1.3           purrr_1.0.2           readr_2.1.4           tidyr_1.3.0          
[13] tibble_3.2.1          ggplot2_3.4.2         tidyverse_2.0.0       OmnipathR_3.9.8      
[17] decoupleR_2.7.1      

loaded via a namespace (and not attached):
  [1] utf8_1.2.3             spatstat.explore_3.2-1 reticulate_1.30       
  [4] R.utils_2.12.2         tidyselect_1.2.0       htmlwidgets_1.6.2     
  [7] Rtsne_0.16             munsell_0.5.0          codetools_0.2-18      
 [10] ica_1.0-3              future_1.33.0          miniUI_0.1.1.1        
 [13] withr_2.5.0            spatstat.random_3.1-5  colorspace_2.1-0      
 [16] progressr_0.13.0       logger_0.2.2           knitr_1.44            
 [19] rstudioapi_0.14        Seurat_4.3.0.9011      stats4_4.1.3          
 [22] ROCR_1.0-11            tensor_1.5             listenv_0.9.0         
 [25] polyclip_1.10-4        bit64_4.0.5            rprojroot_2.0.3       
 [28] parallelly_1.36.0      vctrs_0.6.3            generics_0.1.3        
 [31] xfun_0.40              timechange_0.2.0       R6_2.5.1              
 [34] doParallel_1.0.17      clue_0.3-64            rex_1.2.1             
 [37] spatstat.utils_3.0-3   promises_1.2.0.1       scales_1.2.1          
 [40] vroom_1.6.3            gtable_0.3.3           globals_0.16.2        
 [43] processx_3.8.1         goftest_1.2-3          rlang_1.1.1           
 [46] cyclocomp_1.1.0        GlobalOptions_0.1.2    splines_4.1.3         
 [49] lazyeval_0.2.2         selectr_0.4-2          spatstat.geom_3.2-2   
 [52] checkmate_2.2.0        yaml_2.3.7             reshape2_1.4.4        
 [55] abind_1.4-5            backports_1.4.1        httpuv_1.6.11         
 [58] tools_4.1.3            ellipsis_0.3.2         RColorBrewer_1.1-3    
 [61] BiocGenerics_0.40.0    ggridges_0.5.4         Rcpp_1.0.11           
 [64] plyr_1.8.8             progress_1.2.2         ps_1.7.5              
 [67] prettyunits_1.2.0      deldir_1.0-9           pbapply_1.7-2         
 [70] GetoptLong_1.0.5       cowplot_1.1.1          S4Vectors_0.32.4      
 [73] zoo_1.8-12             SeuratObject_4.1.3     ggrepel_0.9.3         
 [76] cluster_2.1.2          magrittr_2.0.3         data.table_1.14.8     
 [79] scattermore_1.2        lmtest_0.9-40          RANN_2.6.1            
 [82] fitdistrplus_1.1-11    R.cache_0.16.0         matrixStats_1.0.0     
 [85] hms_1.1.3              patchwork_1.1.2        mime_0.12             
 [88] evaluate_0.21          xtable_1.8-4           readxl_1.4.3          
 [91] IRanges_2.28.0         gridExtra_2.3          shape_1.4.6           
 [94] compiler_4.1.3         KernSmooth_2.23-20     crayon_1.5.2          
 [97] R.oo_1.25.0            htmltools_0.5.6        later_1.3.1           
[100] tzdb_0.4.0             MASS_7.3-55            rappdirs_0.3.3        
[103] Matrix_1.5-4           cli_3.6.1              R.methodsS3_1.8.2     
[106] parallel_4.1.3         igraph_1.5.1           pkgconfig_2.0.3       
[109] sp_2.0-0               plotly_4.10.2          spatstat.sparse_3.0-2 
[112] xml2_1.3.5             foreach_1.5.2          rvest_1.0.3           
[115] callr_3.7.3            digest_0.6.33          sctransform_0.3.5.9006
[118] RcppAnnoy_0.0.21       spatstat.data_3.0-1    rmarkdown_2.25        
[121] cellranger_1.1.0       leiden_0.4.3           uwot_0.1.16           
[124] curl_5.0.2             shiny_1.7.4.1          rjson_0.2.21          
[127] lifecycle_1.0.3        nlme_3.1-155           jsonlite_1.8.7        
[130] desc_1.4.2             viridisLite_0.4.2      fansi_1.0.4           
[133] pillar_1.9.0           lattice_0.20-45        fastmap_1.1.1         
[136] httr_1.4.7             survival_3.3-1         glue_1.6.2            
[139] remotes_2.4.2          png_0.1-8              iterators_1.0.14      
[142] bit_4.0.5              stringi_1.7.12         irlba_2.3.5.1         
[145] future.apply_1.11.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "biological_activity",
  "03_pathway_activity.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "biological_activity",
  "03_pathway_activity.Rmd"
))
```
