---
title: "02_jaccard_index_geo"
author: "Tabea M. Soelter"
date: '2023-07-07'
output: html_document
---
**Calculating Jaccard Similarity Index (JI) for CCC genes across cell types**
__Goal__: Calculate similarity, using the Jaccard Index, of identified ligands, receptors, and target genes across cell types, in order to identify cell-type-specific and cell-type-agnostic CCC mechanisms.
  
__Reproducibility__: 
* GitHub: lasseignelab/230313_TS_CCCinHumanAD
* Docker: tsoelter/rstudio_ccc_ad
    * Version: 1.0.3
* HPC: Yes
  * Resources: long partition (150hrs), 8 CPUs, 65GB per CPU

__Data__:
* Name: geo_multinichenet_output.rds
* Location: /data/user/tsoelter/projects/tms_ccc/data/ccc/
* Type: large list
* Processed: Yes, output from MultiNicheNet
 
__Analysis Plan__:
* Load necessary packages 
* Load data 
* Prepare input for Jaccard Index calculation
  * Create dfs with 2 columns: gene, cell type
    * gene refers to either the ligands, receptors, or target genes
* Calculate Jaccard Index
  * Do this for both Ex and In Neurons
  * Dfs should be filtered for condition of interest (AD) and receiver cell type

__Analysis__:
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggcorrplot)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCinAD_tms.R"))

# Setting locale
options(encoding = "UTF-8")
```

# Load data
```{r}
output <- readRDS(here("data", "ccc", "geo_multinichenet_output.rds"))
```

# Set filtering parameters
```{r}
# Number of targets to be used
top_n_target <- 250

# Condition of interest
group_oi <- "AD"

# Receiver cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")

# Sender cell types of interest
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# Table of comparisons of interest
contrast_tbl <- tibble(
  contrast = c("AD-CTRL", "CTRL-AD"),
  group = c("AD", "CTRL")
)
```

# Filter MultiNicheNet output
```{r warning=FALSE}
# Filter by ligand activity and regulatory potential (pearson & spearman corr)
lrt_cor_filtered <- filter_nichenet(output)

# Rename receiver cell types
lrt_cor_filtered$receiver <- ifelse(
  lrt_cor_filtered$receiver == "Inhibitory.Neurons",
  "Inhibitory Neurons",
  "Excitatory Neurons"
)

# Check receivers were renamed and number
table(lrt_cor_filtered$receiver) # Ex 582 In 752
```

# Calculate JI (Between senders)
```{r}
# set receivers again, since I modified them
receivers_oi <- c("Excitatory Neurons", "Inhibitory Neurons")

# ligands
jaccard_results_ligands <- calculate_jaccard(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "ligands"
)

# receptors
jaccard_results_receptors <- calculate_jaccard(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "receptors"
)

# targets
jaccard_results_targets <- calculate_jaccard(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "targets"
)
```

# Calculate JI (Between receivers)
```{r warning=FALSE}
jaccard_receptors_receiver <- calculate_jaccard2(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "receptor"
)

jaccard_targets_receiver <- calculate_jaccard2(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "target"
)
```

# Arrange JI outputs for plotting
```{r}
# receptors
jaccard_receptors <- data.frame(
  cell_types = names(jaccard_receptors_receiver),
  jaccard_index = unlist(jaccard_receptors_receiver),
  stringsAsFactors = FALSE
) %>%
  separate(cell_types,
    into = c(
      "sender",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

rownames(jaccard_receptors) <- NULL

# targets
jaccard_targets <- data.frame(
  cell_types = names(jaccard_targets_receiver),
  jaccard_index = unlist(jaccard_targets_receiver),
  stringsAsFactors = FALSE
) %>%
  separate(cell_types,
    into = c(
      "sender",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

rownames(jaccard_targets) <- NULL

# combined
jaccard_receptors$type <- "Receptors"
jaccard_targets$type <- "Targets"

jaccard_all <- rbind(jaccard_receptors, jaccard_targets)
```

# Plotting
* Lollipop plots of JI between excitatory and inhibitory neurons for receptors and targets
* Combined plot of the above
```{r}
png(here(
  "results",
  "intermediate_outputs",
  "geo",
  "ccc",
  "receptors_lollipop_neurons.png"
))
jaccard_receptors %>%
  arrange(jaccard_index) %>%
  mutate(sender = factor(sender, levels = sender)) %>%
  ggplot(aes(x = sender, y = jaccard_index)) +
  geom_segment(aes(xend = sender, yend = 0)) +
  geom_point(size = 4, color = "#6a3d9a") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_blank()
  ) +
  xlab("Sender Cells") +
  ylab("JI") +
  ylim(0, 0.5) +
  ggtitle("Receptors")
dev.off()

png(here(
  "results",
  "intermediate_outputs",
  "geo",
  "ccc",
  "targets_lollipop_neurons.png"
))
jaccard_targets %>%
  arrange(jaccard_index) %>%
  mutate(sender = factor(sender, levels = sender)) %>%
  ggplot(aes(x = sender, y = jaccard_index)) +
  geom_segment(aes(xend = sender, yend = 0)) +
  geom_point(size = 4, color = "#cab2d6") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_blank()
  ) +
  xlab("Sender Cells") +
  ylab("JI") +
  ylim(0, 0.5) +
  ggtitle("Targets")
dev.off()

png(here("results", "final_outputs", "geo", "combined_lollipop_neurons.png"))
jaccard_all %>%
  ggplot() +
  geom_linerange(aes(x = sender, ymin = 0, ymax = jaccard_index, colour = type),
    position = position_dodge(width = 0.2), size = 1
  ) +
  geom_point(aes(x = sender, y = jaccard_index, colour = type),
    position = position_dodge(width = 0.2), size = 2
  ) +
  coord_flip() +
  scale_color_manual(
    values = c("Receptors" = "salmon", "Targets" = "tomato4"),
    name = "Type"
  ) +
  theme_bw() +
  labs(x = "Sender", y = "JI") +
  ylim(0, 0.5) +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  ) +
  ggtitle("JI between Excitatory and Inhibitory Neurons")
dev.off()
```

# Saving outputs
```{r}
saveRDS(
  jaccard_results_ligands,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_ligands_senders.rds"
  )
)

saveRDS(
  jaccard_results_receptors,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_receptors_senders.rds"
  )
)

saveRDS(
  jaccard_results_targets,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_targets_senders.rds"
  )
)

saveRDS(
  jaccard_receptors,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_receptors_receivers.rds"
  )
)

saveRDS(
  jaccard_targets,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_targets_receivers.rds"
  )
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.1.3 (2022-03-10)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so

locale:
[1] C

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2      styler_1.9.1     here_1.0.1       ggcorrplot_0.1.4 lubridate_1.9.2 
 [6] forcats_1.0.0    stringr_1.5.0    dplyr_1.1.2      purrr_1.0.1      readr_2.1.4     
[11] tidyr_1.3.0      tibble_3.2.1     ggplot2_3.4.2    tidyverse_2.0.0 

loaded via a namespace (and not attached):
  [1] utf8_1.2.3                  spatstat.explore_3.2-1      reticulate_1.30            
  [4] R.utils_2.12.2              tidyselect_1.2.0            htmlwidgets_1.6.2          
  [7] grid_4.1.3                  BiocParallel_1.28.3         Rtsne_0.16                 
 [10] DropletUtils_1.14.2         munsell_0.5.0               codetools_0.2-18           
 [13] ica_1.0-3                   future_1.33.0               miniUI_0.1.1.1             
 [16] withr_2.5.0                 spatstat.random_3.1-5       colorspace_2.1-0           
 [19] progressr_0.13.0            Biobase_2.54.0              knitr_1.43                 
 [22] rstudioapi_0.14             Seurat_4.3.0.9011           stats4_4.1.3               
 [25] SingleCellExperiment_1.16.0 ROCR_1.0-11                 tensor_1.5                 
 [28] listenv_0.9.0               MatrixGenerics_1.6.0        labeling_0.4.2             
 [31] GenomeInfoDbData_1.2.7      polyclip_1.10-4             farver_2.1.1               
 [34] rhdf5_2.38.1                rprojroot_2.0.3             parallelly_1.36.0          
 [37] vctrs_0.6.3                 generics_0.1.3              xfun_0.39                  
 [40] timechange_0.2.0            R6_2.5.1                    GenomeInfoDb_1.30.1        
 [43] locfit_1.5-9.8              rex_1.2.1                   bitops_1.0-7               
 [46] rhdf5filters_1.6.0          spatstat.utils_3.0-3        DelayedArray_0.20.0        
 [49] promises_1.2.0.1            scales_1.2.1                gtable_0.3.3               
 [52] beachmat_2.10.0             globals_0.16.2              processx_3.8.1             
 [55] goftest_1.2-3               rlang_1.1.1                 cyclocomp_1.1.0            
 [58] splines_4.1.3               lazyeval_0.2.2              spatstat.geom_3.2-2        
 [61] yaml_2.3.7                  reshape2_1.4.4              abind_1.4-5                
 [64] httpuv_1.6.11               tools_4.1.3                 ellipsis_0.3.2             
 [67] RColorBrewer_1.1-3          BiocGenerics_0.40.0         ggridges_0.5.4             
 [70] Rcpp_1.0.11                 plyr_1.8.8                  sparseMatrixStats_1.6.0    
 [73] zlibbioc_1.40.0             RCurl_1.98-1.12             ps_1.7.5                   
 [76] deldir_1.0-9                pbapply_1.7-2               cowplot_1.1.1              
 [79] S4Vectors_0.32.4            zoo_1.8-12                  SeuratObject_4.1.3         
 [82] SummarizedExperiment_1.24.0 ggrepel_0.9.3               cluster_2.1.2              
 [85] magrittr_2.0.3              data.table_1.14.8           scattermore_1.2            
 [88] lmtest_0.9-40               RANN_2.6.1                  fitdistrplus_1.1-11        
 [91] R.cache_0.16.0              matrixStats_1.0.0           hms_1.1.3                  
 [94] patchwork_1.1.2             mime_0.12                   evaluate_0.21              
 [97] xtable_1.8-4                IRanges_2.28.0              gridExtra_2.3              
[100] compiler_4.1.3              KernSmooth_2.23-20          crayon_1.5.2               
[103] R.oo_1.25.0                 htmltools_0.5.5             later_1.3.1                
[106] tzdb_0.4.0                  corrplot_0.92               MASS_7.3-55                
[109] Matrix_1.5-4                cli_3.6.1                   R.methodsS3_1.8.2          
[112] parallel_4.1.3              igraph_1.5.0                GenomicRanges_1.46.1       
[115] pkgconfig_2.0.3             sp_2.0-0                    plotly_4.10.2              
[118] scuttle_1.4.0               spatstat.sparse_3.0-2       xml2_1.3.3                 
[121] dqrng_0.3.0                 XVector_0.34.0              callr_3.7.3                
[124] digest_0.6.33               sctransform_0.3.5.9006      RcppAnnoy_0.0.21           
[127] spatstat.data_3.0-1         rmarkdown_2.23              leiden_0.4.3               
[130] uwot_0.1.16                 edgeR_3.36.0                DelayedMatrixStats_1.16.0  
[133] shiny_1.7.4.1               lifecycle_1.0.3             nlme_3.1-155               
[136] jsonlite_1.8.7              Rhdf5lib_1.16.0             desc_1.4.2                 
[139] viridisLite_0.4.2           limma_3.50.3                fansi_1.0.4                
[142] pillar_1.9.0                lattice_0.20-45             fastmap_1.1.1              
[145] httr_1.4.6                  survival_3.3-1              glue_1.6.2                 
[148] remotes_2.4.2               png_0.1-8                   stringi_1.7.12             
[151] HDF5Array_1.22.1            irlba_2.3.5.1               future.apply_1.11.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "ccc",
  "02_jaccard_index_geo.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "ccc",
  "02_jaccard_index_geo.Rmd"
))
```
