---
title: "02_jaccard_index_geo"
author: "Tabea M. Soelter"
date: '2023-07-07'
output: html_document
---
**Calculating Jaccard Similarity Index (JI) for CCC genes across cell types**
__Goal__: Calculate similarity, using the Jaccard Index, of identified ligands, receptors, and target genes across cell types, in order to identify cell-type-specific and cell-type-agnostic CCC mechanisms.
  
__Reproducibility__: 
* GitHub: lasseignelab/230313_TS_CCCinHumanAD
* Docker: tsoelter/rstudio_ccc_ad
    * Version: 1.0.3
* HPC: Yes
  * Resources: long partition (150hrs), 8 CPUs, 65GB per CPU

__Data__:
* Name: geo_multinichenet_output.rds
* Location: /data/user/tsoelter/projects/tms_ccc/data/ccc/
* Type: large list
* Processed: Yes, output from MultiNicheNet
 
__Analysis Plan__:
* Load necessary packages 
* Load data 
* Prepare input for Jaccard Index calculation
  * Create dfs with 2 columns: gene, cell type
    * gene refers to either the ligands, receptors, or target genes
* Calculate Jaccard Index
  * Do this for both Ex and In Neurons
  * Dfs should be filtered for condition of interest (AD) and receiver cell type

__Analysis__:
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggcorrplot)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCinAD_tms.R"))

# Setting locale
options(encoding = "UTF-8")
```

# Load data
```{r}
output <- readRDS(here("data", "ccc", "geo_multinichenet_output.rds"))
```

# Set filtering parameters
```{r}
# Number of targets to be used
top_n_target <- 250

# Receiver cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")

# Sender cell types of interest
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# Table of comparisons of interest
contrast_tbl <- tibble(
  contrast = c("AD-CTRL", "CTRL-AD"),
  group = c("AD", "CTRL")
)
```

# Filter MultiNicheNet output
* Filter by ligand actvity and regulatory potential (person and spearman correlation coefficients)
```{r warning=FALSE}
# filter for AD
group_oi <- "AD"
lrt_cor_filtered_1 <- filter_nichenet(output)

# filter for CTRL
group_oi <- "CTRL"
lrt_cor_filtered_2 <- filter_nichenet(output)

# combine AD and CTRL
lrt_cor_filtered <- rbind(lrt_cor_filtered_1, lrt_cor_filtered_2)

# Rename receiver cell types
lrt_cor_filtered$receiver <- ifelse(
  lrt_cor_filtered$receiver == "Inhibitory.Neurons",
  "Inhibitory Neurons",
  "Excitatory Neurons"
)

# Check receivers were renamed and number
table(lrt_cor_filtered$receiver) # Ex 778 In 840
```

# Calculate JI (Between senders)
```{r}
# set receivers again, since I modified them
receivers_oi <- c("Excitatory Neurons", "Inhibitory Neurons")

# ligands
jaccard_results_ligands <- calculate_jaccard(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "ligands"
)

# receptors
jaccard_results_receptors <- calculate_jaccard(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "receptors"
)

# targets
jaccard_results_targets <- calculate_jaccard(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "targets"
)
```

# Calculate JI (Between receivers)
```{r warning=FALSE}
jaccard_receptors_receiver <- calculate_jaccard2(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "receptor"
)

jaccard_targets_receiver <- calculate_jaccard2(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "target"
)
```

# Arrange JI outputs for plotting
```{r}
# receptors
jaccard_receptors <- data.frame(
  cell_types = names(jaccard_receptors_receiver),
  jaccard_index = unlist(jaccard_receptors_receiver),
  stringsAsFactors = FALSE
) %>%
  separate(cell_types,
    into = c(
      "sender",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

rownames(jaccard_receptors) <- NULL

# targets
jaccard_targets <- data.frame(
  cell_types = names(jaccard_targets_receiver),
  jaccard_index = unlist(jaccard_targets_receiver),
  stringsAsFactors = FALSE
) %>%
  separate(cell_types,
    into = c(
      "sender",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

rownames(jaccard_targets) <- NULL

# combined
jaccard_receptors$type <- "Receptors"
jaccard_targets$type <- "Targets"

jaccard_all <- rbind(jaccard_receptors, jaccard_targets)
```

# Plotting
* Lollipop plots of JI between excitatory and inhibitory neurons for receptors and targets
* Combined plot of the above
```{r}
png(here(
  "results",
  "intermediate_outputs",
  "geo",
  "ccc",
  "receptors_lollipop_neurons.png"
))
jaccard_receptors %>%
  arrange(jaccard_index) %>%
  mutate(sender = factor(sender, levels = sender)) %>%
  ggplot(aes(x = sender, y = jaccard_index)) +
  geom_segment(aes(xend = sender, yend = 0)) +
  geom_point(size = 4, color = "#6a3d9a") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_blank()
  ) +
  xlab("Sender Cells") +
  ylab("JI") +
  ylim(0, 0.5) +
  ggtitle("Receptors")
dev.off()

png(here(
  "results",
  "intermediate_outputs",
  "geo",
  "ccc",
  "targets_lollipop_neurons.png"
))
jaccard_targets %>%
  arrange(jaccard_index) %>%
  mutate(sender = factor(sender, levels = sender)) %>%
  ggplot(aes(x = sender, y = jaccard_index)) +
  geom_segment(aes(xend = sender, yend = 0)) +
  geom_point(size = 4, color = "#cab2d6") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_blank()
  ) +
  xlab("Sender Cells") +
  ylab("JI") +
  ylim(0, 0.5) +
  ggtitle("Targets")
dev.off()

png(here("results", "final_outputs", "geo", "combined_lollipop_neurons.png"))
jaccard_all %>%
  ggplot() +
  geom_linerange(aes(x = sender, ymin = 0, ymax = jaccard_index, colour = type),
    position = position_dodge(width = 0.2), size = 1
  ) +
  geom_point(aes(x = sender, y = jaccard_index, colour = type),
    position = position_dodge(width = 0.2), size = 2
  ) +
  coord_flip() +
  scale_color_manual(
    values = c("Receptors" = "salmon", "Targets" = "tomato4"),
    name = "Type"
  ) +
  theme_bw() +
  labs(x = "Sender", y = "JI") +
  ylim(0, 0.5) +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  ) +
  ggtitle("JI between Excitatory and Inhibitory Neurons")
dev.off()
```

# Saving outputs
```{r}
saveRDS(
  jaccard_results_ligands,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_ligands_senders.rds"
  )
)

saveRDS(
  jaccard_results_receptors,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_receptors_senders.rds"
  )
)

saveRDS(
  jaccard_results_targets,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_targets_senders.rds"
  )
)

saveRDS(
  jaccard_receptors,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_receptors_receivers.rds"
  )
)

saveRDS(
  jaccard_targets,
  here(
    "results",
    "intermediate_outputs",
    "geo",
    "ccc",
    "jaccard_targets_receivers.rds"
  )
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.1.3 (2022-03-10)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so

locale:
[1] C

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2      styler_1.9.1     here_1.0.1       ggcorrplot_0.1.4 lubridate_1.9.2  forcats_1.0.0   
 [7] stringr_1.5.0    dplyr_1.1.3      purrr_1.0.2      readr_2.1.4      tidyr_1.3.0      tibble_3.2.1    
[13] ggplot2_3.4.2    tidyverse_2.0.0 

loaded via a namespace (and not attached):
 [1] pillar_1.9.0      compiler_4.1.3    remotes_2.4.2     R.utils_2.12.2    R.methodsS3_1.8.2 tools_4.1.3      
 [7] digest_0.6.33     R.cache_0.16.0    lifecycle_1.0.3   gtable_0.3.3      timechange_0.2.0  pkgconfig_2.0.3  
[13] rlang_1.1.1       rex_1.2.1         cli_3.6.1         rstudioapi_0.14   xfun_0.40         cyclocomp_1.1.0  
[19] xml2_1.3.5        withr_2.5.0       knitr_1.44        desc_1.4.2        generics_0.1.3    vctrs_0.6.3      
[25] hms_1.1.3         rprojroot_2.0.3   grid_4.1.3        tidyselect_1.2.0  glue_1.6.2        R6_2.5.1         
[31] processx_3.8.1    fansi_1.0.4       farver_2.1.1      callr_3.7.3       tzdb_0.4.0        magrittr_2.0.3   
[37] ps_1.7.5          scales_1.2.1      colorspace_2.1-0  labeling_0.4.2    utf8_1.2.3        stringi_1.7.12   
[43] lazyeval_0.2.2    munsell_0.5.0     crayon_1.5.2      R.oo_1.25.0 

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: < 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "ccc",
  "02_jaccard_index_geo.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "ccc",
  "02_jaccard_index_geo.Rmd"
))
```
