---
title: "06_target_pathways"
author: "Tabea M. Soelter"
date: '2023-05-16'
output: html_document
---
**Pathway analysis of predicted target genes from LIANA and differential NicheNet**
__Goal__: Perform pathway analysis on target genes identified through NicheNet. These genes are potential targets of prioritized ligand-receptor pairs identified through applying the LIANA and differential NicheNet pipeline.
  
__Reproducibility__: 
* GitHub: lasseignelab/230313_TS_CCCinHumanAD
* Docker: tsoelter/rstudio_ccc_ad
    * Version: 1.0.1
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 65GB per CPU

#__Data__: EDIT THIS
* Name: LIANA_NicheNet_ouput.rds
* Location: /data/user/tsoelter/projects/tms_ccc/data/
* Type: tibble
* Processed: Yes, output from LIANA-NicheNet pipeline
 
#__Analysis Plan__: EDIT THIS
* Load necessary packages 
* Load data 
* Prepare input for pathway analysis
  * Pulling out predicted target genes in AD and control neurons
* Determine pathways associated with predicted target genes
  * package: gprofiler2

__Analysis__:
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(gprofiler2)
  library(ggplot2)
  library(lattice)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCinAD_tms.R"))
```

# Load data
* Differential NicheNet outputs
```{r}
# excitatory neurons
vine_cortex_output_ex <- readRDS(here(
  "results",
  "intermediate_outputs",
  "vine_cortex",
  "ccc",
  "ex_neurons_nichenet_output.rds"
))

vine_output_ex <- readRDS(here(
  "results",
  "intermediate_outputs",
  "vine",
  "ccc",
  "ex_neurons_nichenet_output.rds"
))

geo_output_ex <- readRDS(here(
  "results",
  "intermediate_outputs",
  "geo",
  "ccc",
  "ex_neurons_nichenet_output.rds"
))

# inhibitory neurons
vine_cortex_output_in <- readRDS(here(
  "results",
  "intermediate_outputs",
  "vine_cortex",
  "ccc",
  "in_neurons_nichenet_output.rds"
))

vine_output_in <- readRDS(here(
  "results",
  "intermediate_outputs",
  "vine",
  "ccc",
  "in_neurons_nichenet_output.rds"
))

geo_output_in <- readRDS(here(
  "results",
  "intermediate_outputs",
  "geo",
  "ccc",
  "in_neurons_nichenet_output.rds"
))
```

# Pathway analysis
* Filter NicheNet output
  * Receiver cell type
* Assign direction of gene expression
  * "up" = expression values > 0 
  * "down" = expression values < 0 
* Pathway analysis using gprofiler2
* Plotting top enriched pathways of up/down regulated target genes
* Do for all datasets (geo, vine, vine cortex)
```{r}
# create lists of output objects by receiver cell type
ex_output_list <- tibble::lst(
  vine_output_ex,
  geo_output_ex,
  vine_cortex_output_ex,
)

in_output_list <- tibble::lst(
  vine_output_in,
  geo_output_in,
  vine_cortex_output_in
)

# pathway analysis by sender cell type
ex_fea_res <- lapply(ex_output_list,
  pathway_analysis,
  receiver_type = "Excitatory Neurons_AD"
)

in_fea_res <- lapply(in_output_list,
  pathway_analysis,
  receiver_type = "Inhibitory Neurons_AD"
)
```

# Plotting top enriched terms
* Plots are generated by dataset and receiving cell type
_geo_
```{r}
bubbleplot(ex_fea_res$geo_output_ex$combined,
  region = "Prefrontal Cortex Excitatory Neurons",
  file_path = "results/intermediate_outputs/geo/ccc/",
  file_name = "ex_bubbleplot_pathways.png"
)


bubbleplot(in_fea_res$geo_output_in$combined,
  region = "Prefrontal Cortex Inhibitory Neurons",
  file_path = "results/intermediate_outputs/geo/ccc/",
  file_name = "in_bubbleplot_pathways.png"
)
```

_vine_
```{r}
bubbleplot(ex_fea_res$vine_output_ex$combined,
  region = "Hippocampus Excitatory Neurons",
  file_path = "results/intermediate_outputs/vine/ccc/",
  file_name = "ex_bubbleplot_pathways.png"
)


bubbleplot(in_fea_res$vine_output_in$combined,
  region = "Hippocampus Inhibitory Neurons",
  file_path = "results/intermediate_outputs/vine/ccc/",
  file_name = "in_bubbleplot_pathways.png"
)
```

_vine cortex_
```{r}
bubbleplot(ex_fea_res$vine_cortex_output_ex$combined,
  region = "Superior Frontal Cortex Excitatory Neurons",
  file_path = "results/intermediate_outputs/vine_cortex/ccc/",
  file_name = "ex_bubbleplot_pathways.png"
)


bubbleplot(in_fea_res$vine_cortex_output_in$combined,
  region = "Superior Frontal Cortex Inhibitory Neurons",
  file_path = "results/intermediate_outputs/vine_cortex/ccc/",
  file_name = "in_bubbleplot_pathways.png"
)
```

# Session Info
```{r}
sessionInfo() # see output below
```
R version 4.1.3 (2022-03-10)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so

locale:
[1] C

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2      styler_1.9.1     here_1.0.1       lattice_0.20-45  gprofiler2_0.2.1 lubridate_1.9.2 
 [7] forcats_1.0.0    stringr_1.5.0    dplyr_1.1.2      purrr_1.0.1      readr_2.1.4      tidyr_1.3.0     
[13] tibble_3.2.1     ggplot2_3.4.2    tidyverse_2.0.0 

loaded via a namespace (and not attached):
  [1] systemfonts_1.0.4      plyr_1.8.8             igraph_1.4.2           lazyeval_0.2.2        
  [5] sp_1.6-0               splines_4.1.3          listenv_0.9.0          scattermore_0.8       
  [9] digest_0.6.31          htmltools_0.5.5        fansi_1.0.4            magrittr_2.0.3        
 [13] tensor_1.5             cluster_2.1.2          ROCR_1.0-11            tzdb_0.3.0            
 [17] remotes_2.4.2          globals_0.16.2         matrixStats_0.63.0     R.utils_2.12.2        
 [21] timechange_0.2.0       spatstat.sparse_3.0-1  colorspace_2.1-0       ggrepel_0.9.3         
 [25] textshaping_0.3.6      xfun_0.39              callr_3.7.3            crayon_1.5.2          
 [29] RCurl_1.98-1.12        jsonlite_1.8.4         progressr_0.13.0       spatstat.data_3.0-1   
 [33] survival_3.3-1         zoo_1.8-12             glue_1.6.2             polyclip_1.10-4       
 [37] gtable_0.3.3           leiden_0.4.3           R.cache_0.16.0         future.apply_1.10.0   
 [41] abind_1.4-5            scales_1.2.1           spatstat.random_3.1-4  miniUI_0.1.1.1        
 [45] Rcpp_1.0.10            viridisLite_0.4.1      xtable_1.8-4           reticulate_1.28       
 [49] htmlwidgets_1.6.2      rex_1.2.1              httr_1.4.5             RColorBrewer_1.1-3    
 [53] ellipsis_0.3.2         Seurat_4.3.0.9002      ica_1.0-3              farver_2.1.1          
 [57] pkgconfig_2.0.3        R.methodsS3_1.8.2      uwot_0.1.14            deldir_1.0-6          
 [61] utf8_1.2.3             labeling_0.4.2         tidyselect_1.2.0       rlang_1.1.0           
 [65] reshape2_1.4.4         later_1.3.0            munsell_0.5.0          tools_4.1.3           
 [69] cli_3.6.1              generics_0.1.3         ggridges_0.5.4         evaluate_0.20         
 [73] fastmap_1.1.1          ragg_1.2.5             yaml_2.3.7             goftest_1.2-3         
 [77] processx_3.8.1         knitr_1.42             fitdistrplus_1.1-11    RANN_2.6.1            
 [81] pbapply_1.7-0          future_1.32.0          nlme_3.1-155           mime_0.12             
 [85] R.oo_1.25.0            xml2_1.3.3             compiler_4.1.3         rstudioapi_0.14       
 [89] plotly_4.10.1          png_0.1-8              spatstat.utils_3.0-2   stringi_1.7.12        
 [93] cyclocomp_1.1.0        ps_1.7.5               desc_1.4.2             Matrix_1.5-4          
 [97] vctrs_0.6.2            pillar_1.9.0           lifecycle_1.0.3        spatstat.geom_3.1-0   
[101] lmtest_0.9-40          RcppAnnoy_0.0.20       bitops_1.0-7           data.table_1.14.8     
[105] cowplot_1.1.1          irlba_2.3.5.1          httpuv_1.6.9           patchwork_1.1.2       
[109] R6_2.5.1               promises_1.2.0.1       KernSmooth_2.23-20     gridExtra_2.3         
[113] parallelly_1.35.0      codetools_0.2-18       MASS_7.3-55            rprojroot_2.0.3       
[117] withr_2.5.0            SeuratObject_4.1.3     sctransform_0.3.5.9002 parallel_4.1.3        
[121] hms_1.1.3              grid_4.1.3             rmarkdown_2.21         Rtsne_0.16            
[125] spatstat.explore_3.1-0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "ccc",
  "06_target_pathways.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "ccc",
  "06_target_pathways.Rmd"
))
```
