---
title: "figure_S8"
author: "Tabea M. Soelter"
date: '2023-10-08'
output: html_document
---
**Plotting Supplemental Figure 8**

__Goal__: This is a script for generating supplemental figure 8 of my manuscript.
  
__Reproducibility__: 
* GitHub: lasseignelab/230313_TS_CCCinHumanAD
* Docker: tsoelter/rstudio_ccc_ad
    * Version: 1.0.5
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 85GB per CPU

__Data__:
*MultiNicheNet outputs*
* Name: geo_multinichenet_output.rds, gse_multinichenet_output.rds
* Location: data/ccc/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * MultiNicheNet outputs
* Plot individual panels
  * A: Alluvial plot of overlapping LRs upregulated in AD colored by sender
  * B: Alluvial plot of overlapping LRs upregulated in CTRL colored by sender
* Compile figure
* Save figure

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(ggalluvial)
  library(reshape2)
  library(ComplexHeatmap)
  library(ggpubr)
  library(cowplot)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCinAD_tms.R"))
```

# Load data
```{r}
gse_output <- readRDS(here("data", "ccc", "gse_multinichenet_output_final.rds"))
geo_output <- readRDS(here("data", "ccc", "geo_multinichenet_output_final.rds"))
```

# Colors
```{r}
colors <- c(
  `Oligodendrocytes` = "royalblue4",
  `Microglia` = "orchid3",
  `OPCs` = "mediumpurple1",
  `Astrocytes` = "cornflowerblue",
  `Excitatory Neurons` = "slateblue3",
  `Inhibitory Neurons` = "darkslategray3",
  `Pericytes` = "deepskyblue3",
  `Endothelial cells` = "darkolivegreen4"
)
```

# Set filtring criteria
```{r}
top_n_target <- 250

receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

contrast_tbl <- tibble(
  contrast = c("AD-CTRL", "CTRL-AD"),
  group = c("AD", "CTRL")
)
```

# Panel A
* Filter MultiNicheNet outputs by cell types of interests and condition
```{r}
group_oi <- "AD"

lr_target_prior_cor_filtered_gse <- filter_nichenet(gse_output)

lr_target_prior_cor_filtered_geo <- filter_nichenet(geo_output)
```

* Overlapping LRs in AD
```{r}
geo_id <- lr_target_prior_cor_filtered_geo$id %>% unique()

gse_id <- lr_target_prior_cor_filtered_gse$id %>% unique()

overlap <- intersect(geo_id, gse_id)

length(overlap)
```

* combine filtered objects
```{r}
geo_ccc <- lr_target_prior_cor_filtered_geo %>%
  dplyr::select(sender, receiver, ligand, target, receptor, id) %>%
  filter(id %in% overlap) %>%
  mutate(origin = "geo")

gse_ccc <- lr_target_prior_cor_filtered_gse %>%
  dplyr::select(sender, receiver, ligand, target, receptor, id) %>%
  filter(id %in% overlap) %>%
  mutate(origin = "gse")

ccc_combined <- rbind(geo_ccc, gse_ccc) %>%
  select(sender, receiver, ligand, receptor, id) %>%
  unique()
```

* plot AD LR alluvial
```{r}
ccc_combined$ligand <- reorder(
  ccc_combined$ligand,
  desc(ccc_combined$sender)
)
ccc_combined$receptor <- reorder(
  ccc_combined$receptor,
  desc(ccc_combined$sender)
)

lr_ad <- ggplot(ccc_combined,
  aes(axis1 = ligand, axis2 = receptor),
  label = stratum
) +
  scale_x_discrete(limits = c("Ligand", "Receptor")) +
  xlab("Type") +
  geom_alluvium() +
  stat_alluvium(lode.guidance = "frontback") +
  geom_stratum(aes(fill = sender)) +
  geom_label(
    stat = "stratum",
    fill = "white",
    aes(label = after_stat(stratum))
  ) +
  geom_flow(aes(fill = sender), stat = "alluvium", color = "black") +
  scale_fill_manual(values = colors) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum)),
    fontface = "bold",
    fontsize = 12
  ) +
  scale_x_discrete(limits = c("Ligand", "Receptor")) +
  theme_void() +
  theme(
    axis.text.x = element_text(vjust = 2, face = "bold", size = 12),
    legend.position = c(0.9, 0.5),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 14)
  ) +
  guides(fill = guide_legend(title = "Sender"))

lr_ad
```

# Panel B
* Filter MultiNicheNet outputs by cell types of interests and condition
```{r}
group_oi <- "CTRL"

lr_target_prior_cor_filtered_gse <- filter_nichenet(gse_output)

lr_target_prior_cor_filtered_geo <- filter_nichenet(geo_output)
```

* Overlapping LRs in CTRL
```{r}
geo_id <- lr_target_prior_cor_filtered_geo$id %>% unique()

gse_id <- lr_target_prior_cor_filtered_gse$id %>% unique()

overlap <- intersect(geo_id, gse_id)

length(overlap)
```

* combine filtered objects
```{r}
geo_ccc <- lr_target_prior_cor_filtered_geo %>%
  dplyr::select(sender, receiver, ligand, target, receptor, id) %>%
  filter(id %in% overlap) %>%
  mutate(origin = "geo")

gse_ccc <- lr_target_prior_cor_filtered_gse %>%
  dplyr::select(sender, receiver, ligand, target, receptor, id) %>%
  filter(id %in% overlap) %>%
  mutate(origin = "gse")

ccc_combined <- rbind(geo_ccc, gse_ccc) %>%
  select(sender, receiver, ligand, receptor, id) %>%
  unique()
```

* plot CTRL LR alluvial
```{r}
ccc_combined$ligand <- reorder(
  ccc_combined$ligand,
  desc(ccc_combined$sender)
)
ccc_combined$receptor <- reorder(
  ccc_combined$receptor,
  desc(ccc_combined$sender)
)

lr_ctrl <- ggplot(ccc_combined,
  aes(axis1 = ligand, axis2 = receptor),
  label = stratum
) +
  scale_x_discrete(limits = c("Ligand", "Receptor")) +
  xlab("Type") +
  geom_alluvium() +
  stat_alluvium(lode.guidance = "frontback") +
  geom_stratum(aes(fill = sender)) +
  geom_label(
    stat = "stratum",
    fill = "white",
    aes(label = after_stat(stratum))
  ) +
  geom_flow(aes(fill = sender), stat = "alluvium", color = "black") +
  scale_fill_manual(values = colors) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum)),
    fontface = "bold",
    fontsize = 12
  ) +
  scale_x_discrete(limits = c("Ligand", "Receptor")) +
  theme_void() +
  theme(
    axis.text.x = element_text(vjust = 2, face = "bold", size = 12),
    legend.position = c(0.9, 0.5),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 14)
  ) +
  guides(fill = guide_legend(title = "Sender"))

lr_ctrl
```

# Compile figure
```{r}
figure_s8 <- ggarrange(lr_ad,
  lr_ctrl,
  labels = c("A", "B"),
  common.legend = TRUE,
  legend = "bottom"
)

figure_s8
```

# Save figure
```{r}
png(
  here(
    "results",
    "figures",
    "supp_figure8.png"
  ),
  height = 250,
  width = 350,
  units = "mm",
  res = 300
)
figure_s8
dev.off()
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.1.3 (2022-03-10)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so

locale:
[1] C

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2           styler_1.9.1          cowplot_1.1.1         ggpubr_0.6.0          ComplexHeatmap_2.10.0
 [6] reshape2_1.4.4        ggalluvial_0.12.5     lubridate_1.9.2       forcats_1.0.0         stringr_1.5.0        
[11] dplyr_1.1.3           purrr_1.0.2           readr_2.1.4           tidyr_1.3.0           tibble_3.2.1         
[16] ggplot2_3.4.2         tidyverse_2.0.0       here_1.0.1           

loaded via a namespace (and not attached):
 [1] matrixStats_1.0.0   doParallel_1.0.17   RColorBrewer_1.1-3  rprojroot_2.0.3     R.cache_0.16.0     
 [6] tools_4.1.3         backports_1.4.1     utf8_1.2.3          R6_2.5.1            lazyeval_0.2.2     
[11] BiocGenerics_0.40.0 colorspace_2.1-0    GetoptLong_1.0.5    withr_2.5.0         tidyselect_1.2.0   
[16] gridExtra_2.3       processx_3.8.1      compiler_4.1.3      cli_3.6.1           xml2_1.3.5         
[21] desc_1.4.2          labeling_0.4.2      scales_1.2.1        callr_3.7.3         digest_0.6.33      
[26] rmarkdown_2.25      R.utils_2.12.2      pkgconfig_2.0.3     htmltools_0.5.6     fastmap_1.1.1      
[31] rlang_1.1.1         GlobalOptions_0.1.2 rstudioapi_0.14     shape_1.4.6         farver_2.1.1       
[36] generics_0.1.3      car_3.1-2           R.oo_1.25.0         magrittr_2.0.3      Rcpp_1.0.11        
[41] munsell_0.5.0       S4Vectors_0.32.4    fansi_1.0.4         abind_1.4-5         lifecycle_1.0.3    
[46] R.methodsS3_1.8.2   stringi_1.7.12      yaml_2.3.7          carData_3.0-5       plyr_1.8.8         
[51] parallel_4.1.3      crayon_1.5.2        circlize_0.4.15     hms_1.1.3           knitr_1.44         
[56] ps_1.7.5            pillar_1.9.0        rjson_0.2.21        ggsignif_0.6.4      codetools_0.2-18   
[61] stats4_4.1.3        glue_1.6.2          evaluate_0.21       rex_1.2.1           remotes_2.4.2      
[66] png_0.1-8           vctrs_0.6.3         tzdb_0.4.0          foreach_1.5.2       gtable_0.3.3       
[71] clue_0.3-64         xfun_0.40           broom_1.0.5         cyclocomp_1.1.0     rstatix_0.7.2      
[76] iterators_1.0.14    IRanges_2.28.0      cluster_2.1.2       timechange_0.2.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: < 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "manuscript_figures",
  "figure_S8.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "manuscript_figures",
  "figure_S8.Rmd"
))
```
