---
title: "figure_2"
author: "Tabea M. Soelter"
date: '2023-05-19'
output: html_document
---
**Plotting Figure 2**

__Goal__: This is a script for generating figure 2 of my manuscript.
  
__Reproducibility__: 
* GitHub: lasseignelab/230313_TS_CCCinHumanAD
* Docker: tsoelter/rstudio_ccc_ad
    * Version: 1.0.5
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 85GB per CPU

__Data__:
*Seurat objects*
* Name: geo_processed_seurat.rds
* Location: data/seurat_preprocessing/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * Seurat object
* Plot individual panels
  * A: UMAP w cell type assignments
  * B: Violin plot with expression of marker genes
  * C: UMAP split by condition
  * D: Split barplot of cell type proportions
* Compile figure
* Save figure

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(circlize)
  library(ComplexHeatmap)
  library(RColorBrewer)
  library(reshape2)
  library(patchwork)
  library(cowplot)
  library(here)
  library(styler)
  library(lintr)
})
```

# Load in object
```{r}
geo_object <- readRDS(here(
  "data",
  "seurat_preprocessing",
  "geo_processed_seurat.rds"
))
```

# Set color palette
```{r}
colors <- c(
  `Oligodendrocytes` = "royalblue4",
  `Microglia` = "orchid3",
  `OPCs` = "mediumpurple1",
  `Astrocytes` = "cornflowerblue",
  `Excitatory Neurons` = "slateblue3",
  `Inhibitory Neurons` = "darkslategray3",
  `Pericytes` = "deepskyblue3",
  `Endothelial cells` = "darkolivegreen4"
)
```

# Figure 2A
* UMAP of assigned cell types
```{r}
umap_1 <- DimPlot(geo_object,
  cols = colors,
  label = TRUE,
  label.box = TRUE,
  label.color = "white",
  label.size = 4,
  repel = TRUE
) +
  theme(legend.text = element_text(face = "bold")) +
  theme(axis.text = element_text(face = "bold")) +
  theme(axis.title = element_text(face = "bold")) +
  NoLegend()

png(
  here(
    "results",
    "final_outputs",
    "geo",
    "UMAP_final2.png"
  ),
  width = 600,
  height = 600
)
umap_1
dev.off()
```

# Figure 2B
* Violin plot of the marker gene expression of the assigned cell types
```{r}
markers <- c(
  "MBP",
  "ST18",
  "QDPR",
  "DOCK8",
  "APBB1IP",
  "ARHGAP15",
  "PCDH15",
  "PDGFRA",
  "ADGRV1",
  "AQP4",
  "GFAP",
  "RALYL",
  "IL1RAPL2",
  "NXPH1",
  "GRIK1",
  "PDGFRB",
  "FLT1",
  "VWF"
)

violin <- VlnPlot(geo_object,
  markers,
  stack = TRUE,
  sort = FALSE,
  flip = TRUE,
  assay = "RNA",
  fill.by = "ident",
  cols = colors
) +
  labs(fill = "Cell Type") +
  theme_minimal() +
  theme(legend.position = "left") +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  theme(panel.grid = element_blank()) +
  theme(strip.text.y.right = element_text(angle = 0, face = "bold")) +
  theme(legend.title = element_text(face = "bold")) +
  theme(axis.title = element_text(face = "bold")) +
  theme(axis.text = element_text(face = "bold")) +
  theme(legend.text = element_text(face = "bold")) +
  ylab("Marker Gene Expression")

png(
  here(
    "results",
    "final_outputs",
    "geo",
    "markers_violin_final.png"
  ),
  width = 500,
  height = 700
)
violin
dev.off()
```

# Figure 2C
* UMAP split by condition
```{r}
umap_2 <- DimPlot(geo_object,
  group.by = "orig.ident",
  cols = "Paired",
  shuffle = TRUE,
  seed = 42
) +
  theme(legend.title = element_text(face = "bold")) +
  labs(color = "Condition") +
  theme(legend.text = element_text(face = "bold")) +
  theme(axis.text = element_text(face = "bold")) +
  theme(axis.title = element_text(face = "bold"))

# remove title
umap_2 <- umap_2 + ggtitle(NULL)

png(
  here(
    "results",
    "final_outputs",
    "geo",
    "UMAP_condition_final.png"
  ),
  width = 620,
  height = 600
)
umap_2
dev.off()
```

# Figure 2D
* Stacked barplot of cell type proportion distribution across conditions
```{r}
#  pull out cell type information
df <- table(Idents(geo_object), geo_object$orig.ident) %>% as.data.frame()

# change classs for plotting purposes
df$Var1 <- as.character(df$Var1)

barplot <- ggplot(df, aes(x = fct_rev(Var1), y = Freq, fill = Var2)) +
  theme_bw(base_size = 15) +
  geom_col(position = "fill", width = 0.5) +
  xlab("Cell Type") +
  ylab("Proportion") +
  scale_fill_manual(values = brewer.pal(12, "Paired")) +
  theme(legend.title = element_text(face = "bold")) +
  theme(axis.title = element_text(face = "bold")) +
  theme(axis.text = element_text(face = "bold")) +
  theme(legend.text = element_text(face = "bold")) +
  coord_flip() +
  labs(fill = "Condition")

png(
  here(
    "results",
    "final_outputs",
    "geo",
    "stacked_barplot_final.png"
  ),
  width = 600,
  height = 900
)
barplot
dev.off()
```

# Compile figure 2
```{r}
fig2 <- plot_grid(umap_1,
  umap_2,
  violin,
  barplot,
  labels = c("A", "C", "B", "D"),
  rel_heights = c(1, 2)
)

png(here("results", "figures", "figure2.png"),
  width = 300,
  height = 300,
  units = "mm",
  res = 300
)
fig2
dev.off()
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.1.3 (2022-03-10)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so

locale:
[1] C

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods  
[8] base     

other attached packages:
 [1] lintr_3.0.2           styler_1.9.1          here_1.0.1           
 [4] cowplot_1.1.1         patchwork_1.1.2       reshape2_1.4.4       
 [7] RColorBrewer_1.1-3    ComplexHeatmap_2.10.0 circlize_0.4.15      
[10] SeuratObject_4.1.3    Seurat_4.3.0.9011     lubridate_1.9.2      
[13] forcats_1.0.0         stringr_1.5.0         dplyr_1.1.3          
[16] purrr_1.0.2           readr_2.1.4           tidyr_1.3.0          
[19] tibble_3.2.1          ggplot2_3.4.2         tidyverse_2.0.0      

loaded via a namespace (and not attached):
  [1] plyr_1.8.8             igraph_1.5.1           lazyeval_0.2.2        
  [4] sp_2.0-0               splines_4.1.3          listenv_0.9.0         
  [7] scattermore_1.2        digest_0.6.33          foreach_1.5.2         
 [10] htmltools_0.5.6        fansi_1.0.4            magrittr_2.0.3        
 [13] tensor_1.5             cluster_2.1.2          doParallel_1.0.17     
 [16] ROCR_1.0-11            remotes_2.4.2          tzdb_0.4.0            
 [19] globals_0.16.2         matrixStats_1.0.0      R.utils_2.12.2        
 [22] timechange_0.2.0       spatstat.sparse_3.0-2  colorspace_2.1-0      
 [25] ggrepel_0.9.3          xfun_0.40              callr_3.7.3           
 [28] crayon_1.5.2           jsonlite_1.8.7         progressr_0.13.0      
 [31] spatstat.data_3.0-1    survival_3.3-1         zoo_1.8-12            
 [34] iterators_1.0.14       glue_1.6.2             polyclip_1.10-4       
 [37] gtable_0.3.3           leiden_0.4.3           GetoptLong_1.0.5      
 [40] R.cache_0.16.0         future.apply_1.11.0    shape_1.4.6           
 [43] BiocGenerics_0.40.0    abind_1.4-5            scales_1.2.1          
 [46] spatstat.random_3.1-5  miniUI_0.1.1.1         Rcpp_1.0.11           
 [49] viridisLite_0.4.2      xtable_1.8-4           clue_0.3-64           
 [52] reticulate_1.30        stats4_4.1.3           htmlwidgets_1.6.2     
 [55] rex_1.2.1              httr_1.4.7             ellipsis_0.3.2        
 [58] ica_1.0-3              farver_2.1.1           pkgconfig_2.0.3       
 [61] R.methodsS3_1.8.2      uwot_0.1.16            deldir_1.0-9          
 [64] utf8_1.2.3             labeling_0.4.2         tidyselect_1.2.0      
 [67] rlang_1.1.1            later_1.3.1            munsell_0.5.0         
 [70] tools_4.1.3            cli_3.6.1              generics_0.1.3        
 [73] ggridges_0.5.4         fastmap_1.1.1          goftest_1.2-3         
 [76] processx_3.8.1         knitr_1.44             fitdistrplus_1.1-11   
 [79] RANN_2.6.1             pbapply_1.7-2          future_1.33.0         
 [82] nlme_3.1-155           mime_0.12              R.oo_1.25.0           
 [85] xml2_1.3.5             compiler_4.1.3         rstudioapi_0.14       
 [88] plotly_4.10.2          png_0.1-8              spatstat.utils_3.0-3  
 [91] stringi_1.7.12         ps_1.7.5               cyclocomp_1.1.0       
 [94] desc_1.4.2             lattice_0.20-45        Matrix_1.5-4          
 [97] vctrs_0.6.3            pillar_1.9.0           lifecycle_1.0.3       
[100] spatstat.geom_3.2-2    lmtest_0.9-40          GlobalOptions_0.1.2   
[103] RcppAnnoy_0.0.21       data.table_1.14.8      irlba_2.3.5.1         
[106] httpuv_1.6.11          R6_2.5.1               promises_1.2.0.1      
[109] KernSmooth_2.23-20     gridExtra_2.3          IRanges_2.28.0        
[112] parallelly_1.36.0      codetools_0.2-18       MASS_7.3-55           
[115] rprojroot_2.0.3        rjson_0.2.21           withr_2.5.0           
[118] sctransform_0.3.5.9006 S4Vectors_0.32.4       parallel_4.1.3        
[121] hms_1.1.3              Rtsne_0.16             spatstat.explore_3.2-1
[124] shiny_1.7.4.1

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: < 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "manuscript_figures",
  "figure_2.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "manuscript_figures",
  "figure_2.Rmd"
))
```
